Towards a Wide-Coverage Grammar for Swedish\\ Using GF

### 2012-01-17 doc/report/report.tex
last report?

\abstract{
This thesis describes work towards a wide-coverage grammar for
parsing and generating Swedish text.
We do this by using the dependently typed grammar formalism GF,
a functional programming language specialized at describing grammars.
The idea is to 
%The goal of this thesis is to enable open-domain parsing for Swedish
%by using the GF \\grammar formalism,
%by %The idea has been to% 
combine existing language resources with
new techniques with an aim to achieve a parser
for unrestricted Swedish.
To reach this goal, problems of computational as well as linguistic nature had
to be solved.  
\\The work includes the development of the grammar --
%in the GF formalism
to identify and formalize 
grammatical constructions %specific to and
frequent in Swedish -- as well as methods for importing a large-scale lexicon and for evaluating
the parser. % have been investigated and formalised in the functional
%programming language.
%~~achieve a covering Swedish~~ grammar implemented
%and the aim to reaching our goal;
%TODO first sentence should be more, no understatements!
%in GF -- to model natural language by functional programming. 
We present the methods and technologies used 
and discuss the advantages and problems of using GF
for modeling large-scale grammars and 
how our long-term goal can be reached 
%the possibilities
%of reaching our long-term goal % -- to accomplish a parser for unrestricted Swedish -- 
by combining our rule-based grammar with statistical means.
%We compare and discuss .
%and for obtaining a system for analysing written text.
\\Our contribution is a wide-coverage GF lexicon, a translation of a Swedish treebank
into the GF notation and an extended Swedish grammar implementation. The grammar is based
on the multilingual abstract syntax given in the GF resource library, and now
also covers constructions specific to Swedish.
We further give an example of the advantage of using dependent types when
describing grammar and syntax, in this case for dealing with reflexive pronouns.
%the linguistic phenomena

@@

\newpage
\pagestyle{empty}
\section*{Acknowledgments}
\vspace{5mm}
Many people have helped me during this project and made work thesis possible.
I would first of all like to thank
Center of Language Technology, that has funded the project.

@@

The use of controlled language thus gives the possibility of good and reliable
translation and GF has successfully been used for this in a number of projects.
%like %Some examples of projects using GF are 
WebAlt \cite{webalt} aims to develop language independent
material for mathematics problems by using GF grammars.
The formal software verification tool
KeY \cite{key} used GF for translating formal specification to
natural language.
%>> key, talk, hats, bom (see references in Ramonas thesis).
%>> use talk here and move key to other place??
GF has further been used for describing dialogue grammars \cite{talk}, and
the framework is one of the main component in the
European project MOLTO \footnote{http://www.molto-project.eu/}, which develop
online translation between to 15 languages.\\

@@

  fun 
    Pred : N -> V -> S ;
    cat_N : N ;
    sleep_V : V ;
 }
\end{verbatim}
\caption{A small example of an abstract syntax}
\label{fig:gfAbstract1}
\end{figure}

@@

We now have two implementations of the abstract, one for Swedish and one for English.
The resulting GF grammar is able both to
parse a string to an abstract tree and to go in the other direction; to produce
a string of natural language given an abstract tree. This step is called linearization.
Translation is a consequence of this, we can parse a Swedish string and then 
linearize the resulting abstract tree to English.

@@

\noindent The semantic aspect of Saldo requires that words with multiple meanings are
separated into different entries.
%TODO change word? uppskatta, känna, få
The word `uppskatta' for example, has two entries, with slightly
different semantics: \emph{enjoy}
\enumsentence{Jag uppskattar teater. \\
``I enjoy theater."}
or \emph{estimate}. 
\enumsentence{Värdet uppskattades till 5 miljoner.\\
``The value was estimated to 5 million."}
%\enumsentence{Barnen växte varje dag\\
%``The children grew for each day"}
%or \emph{vegetate}. 
%\enumsentence{Det växte en gammal ek där.\\
%``There stood an old oak."}
`uppskatta' is however declined the same way for both interpretations of the word.
Hence \textit{Saldo's morphological lexicon} only keeps one entry of `uppskatta'.
For the homonyms that do not share the whole inflection tables, different entries are 
kept also in the morphological lexicon.\\
For our purpose, only the morphological section, provided under LPGL in XML
format, was needed.
The data can be processed and translated to GF format as described in 
section \ref{sec:prog.saldo}.

@@

%
%There are a number of differences between and saldo Till skillnad från GF,
%saldo does not contain forms that aren't used gf generates all forms. A form
%needed to generate the gf table may not be given in saldo such as singular
%forms for 'byxor'. Also contains idioms, 'hålla huvudet kallt' and has
%different grammatical analyse, a lot of pronouns. Example. 
Most but not all words of these categories have been imported.
One reason why the importing phase would fail 
is that Saldo, unlike GF, only contains the actually used word forms.
% does not contain information about ~~word forms that are not used~~.
For technical reasons, the smart paradigm might need forms never used.
%for this is Deponent verbs
%are for example given with their (often non-existing) active form, al
%In some cases those forms are needed by the smart paradigm. 
Consider for example
the plural tantum % \cite[44]{SAG} 
noun \emph{`glasögon'} (\emph{`glasses'}).
The smart paradigm requires a singular form, and since the program could not
find this in Saldo, there was no way of adding the lemma to the lexicon. 
When the program failed to import a noun, this was often the explanation.
%bbMany nouns fai, \emph{`kläder'} (\emph{`clothes'} etc.
Words of this type may be added manually, for \emph{`glasögon'} we could use
the ostensibly correct singular form `glasöga', although this
has another meaning (`glass-eye').
The same problem occurred for the irregular s-verbs,
(\emph{`synas'} \emph{(`show')} or \emph{umgås} \emph{(`socialize')})
which made up 61.5 \% of the failing verbs of type \verb_vb_.\\
%`frodas' (`flourish'), `lyckas' (`succeeed').
In a few cases the smart paradigms could not generate the correct declination.\\

@@

%       The table suggests that many 
%       >> 2537 without IDs and only the ones built of Alpha,only 552 without compounds, still contains abbrev.
%       >> 494 without abbr. spelling errors
%       >> 142 more then one time.
%       numerals and proper names, but some names are still there). In addition, there were 1817 words
%       >> 223 were IDS.++146 HS.
%       >> 426 more than one time, with compounds.
%       >> give reasons! should look good, not bad!
A list of words that were given different labels in GF than in Talbanken has been
composed, consisting of about 1600 entries. Many of those are
acceptable and reflects the difference 
made in the analyses, such as the examples in table \ref{tab:saldodiff}.
Others are examples of words that are still missing from
the lexicon.
\begin{figure}[h]
\begin{tabular}{|lll|}
\hline
Word    & Talbanken tag & GF category \\
\hline
måste   & MVPS          & VV \\
allting & POTP          & N \\
få      & POZP          & Det \\
\hline
\end{tabular}
\caption{}
\label{tab:saldodiff}
\end{figure}

%Since Saldo does not give any valency information, neither does the imported lexicon.
%The lexicon does not contain any information.
Valency information, which is crucial for GF, is not given in Saldo and
hence not in the imported lexicon. It remains as future work to find methods 
%using the results from the mapping (see section \ref{sec:Mapping})
to extract this information from Talbanken and to automatically build
it into the lexicon. 
%The saddest thing: no valency info. Crucial for GF. Only reflexive and verbs with particles
%have any information. The earlier explained techniques may be used for this in Talbanken.
%To get even bigger material, use Korp, but then no guarantee.

@@

\begin{figure}[h]
\begin{verbatim}
DetCN : Det                  -> CN     -> NP ;
        indefinite, singular -> flicka -> "en flicka"
        definitive, singular -> flicka -> "flickan"
\end{verbatim}        
\caption{}
\label{gfcode:DetCN}
\end{figure}
\newpage
%The two basic articles are \verb-IndefArt- and \verb-DefArt- which
%uses the post-modified form of the noun: katten.
Some rules for determination were described in section \ref{sec:swedishnoun};
if we have a common noun phrase consisting of the parts \emph{`liten'}
(\emph{`small'}) and \emph{`katt'} (\emph{`cat'}), there are three ways they
can be combined, as shown in figure \ref{tab:detdef}. The noun may be used in
definite or indefinite form and the adjective in its weak or strong form \cite[p. 31]{gunlog}.
\begin{figure}[h]
\begin{tabular}{|lll|l|}
\hline
\textbf{Determiner} & \textbf{Adjective} & \textbf{Noun} & \textbf{DetSpecies}\\
\hline
en  &  liten $[${\sc -Def}$]$& katt   $[${\sc -Det}$]$& \textbf{ DIndef}\\
min &  lilla $[${\sc +Def}$]$& katt   $[${\sc -Det}$]$  & \textbf{ DDef Indef} \\
den &  lilla $[${\sc +Def}$]$& katten $[${\sc +Det}$]$& \textbf{ DDef Def} \\
\hline
\end{tabular}        
\caption{}\label{tab:detdef}
\end{figure}

@@

\noindent Swedish verbs may take up to five arguments \cite[p. 53]{stymne}.
These may be prepositions, particles, reflexive object, indirect objects
and direct objects. 
\enumsentence{\shortex{7}
{Jag &tar& med &mig& den& till& honom}
{\emph{I} &\emph{take}&\emph{with} &\emph{me}& \emph{it}& \emph{to}& \emph{him}}
{``I bring it to him"}
\label{sent:tamed}}
The verb `ta' in sentence (\ref{sent:tamed}) takes one particle (\emph{`med'}),
one preposition (\emph{`till'}) and two objects: \emph{`den'} and \emph{`honom'}.
In a GF lexicon this verb is given the
category \verb-V3-, %classifing it as 
a \textit{three-place verb}, taking two objects.
The notion \verb-V3- is motivated by the formal translation: \verb-bring(I,it,him)-.\\
%these will be encoded in different ways.
Particles are given in the lexicon, as well as the prepositions that are
chosen by the verb, since these may not vary.
The entry for `\emph{ta med}', as used in sentence (\ref{sent:tamed}), is
described as follows:
%, assuming that we already have an constant \verb-take_V- which produces the correct conjugation for `ta':
\begin{verbatim}
ta_med_V3 = dirV3 (reflV (partV (take_V "med"))) (mkPrep "till") ;
\end{verbatim}
%The category \verb-V3- tells us that the verb takes two objects, % (`den' and `honom').
The function \verb-dirV3- creates a three-place verb, where the first object is direct and
the second is to be used with the preposition given as the last argument: `\emph{till}'.
\verb-reflV- shows that the verb always is used with a reflexive
pronoun and \verb-partV- gives the particle \emph{`med'}.
The fact that the chosen prepositions is attached to the verb in the lexicon
causes the parse tree visualization algorithm to group them together.
This is also the case for particles, cf. parse tree \ref{pic:tittapart}
and \ref{pic:tittapa}.
\begin{figure}[h]
\subfloat[Verb with a chosen preposition]{\label{pic:tittapa}\includegraphics[height=70mm]{tittapa.jpg}}
\hspace{10mm}
\subfloat[Verb with particle]{\label{pic:tittapart}\includegraphics[height=76mm]{tittapart.pdf}}
\caption{The visualized parse trees do not show the internal difference of chosen prepositions and particles}
\label{fig:translationtrees}
\end{figure}
\newpage
As already stated, the visualized parse trees is not a complete representation,
even if the verb phrases in the visualizations look the same, the two cases
are treated and represented differently internally. The fronting of the
preposition, as in of sentence (\ref{sent:tittapa}a)., is accepted but
fronting of particles, as in \ref{sent:tittapa}b., is not.
%TODO use png to get correct size
\eenumsentence{
\item[a]\shortex{4}
{På &pojken& tittar& du.}
{\emph{on} &\emph{the boy}& \emph{look}& \emph{you}}
{``You look at the boy."}
\item[b]\shortexnt{6}
{*På& när& du& springer& tittar& jag}
{\emph{~on}& \emph{when}& \emph{you}& \emph{run}& \emph{watch}& \emph{I}}
\label{sent:tittapa}}

@@

Like verb phrases, a clause may also be missing an object and then has the type
\verb-ClSlash-.
The \verb-ClSlash- is formed by a \verb-VPSlash- which is given a subject.
This is a convenient way to form questions, relative clauses
and topicalized clauses (see figure \ref{fig:clause2}), as introduced in \cite{gazdar}.\\

@@

%What was in the resources
%description of the existing grammar, the types, the possibilities
%VP, NP.
%valency built in (problems with ställa - på, under, vid).
%Variants, nonexist.
%coverage of standard Swedish.

\section{Development of the grammar}
\label{sec:Added}
%The resource grammar is meant to cover the most important grammatical structures
%present in many languages, but not the language specific constructions.
It has earlier been hard to identify missing constructions of the Swedish
implementation, since there was no large resource available to evaluate it on.

@@

%'han såg sin mammas bil', they same could be applied to
%APs for example \ref{ex:ReflAP} or Adv for
%\enumsentence{Han leker oftare än sin kompis}
% (ComparAdvAdj)
%Atm you can't focus this kind of object, then a ClSlash need info about the
%subject. FunRP also tricky (and what does it mean, we can say
%Han såg sin katt i vilken musen låg), in this case RP needs to give more information.
%Kort sagt, the information about the subject must be spread
%in different parts of the sentence. 
%Could also decide to always produc 'sin' -> jag såg sin bil
%to avoid the dependence, but not nice. Or keep a flag in NP saying if it is
%in third person, and add nonexist for others. But this way we can also get rid
%of ReflVP and the possessive prons \verb|PossPron| should only be used for somebody else:
%'jag såg din bil'.
%Unfortunaly, have to change ReflVP, this only allows this in object for VPSlash, not FocObj,
%not 'alla sina', usw. Want it to handle conjuction as in example \ref{ex:refGen}
%etc.
%varandra, works the same but has no case for Sg.
%%

\subsection{A second future tense: ``Kommer att"}
Swedish and Norwegian use two auxiliary verbs for future tense \\
\cite[p. 246]{H&H}:
%\cite[Tempus \textsection 28]{SAG}.
\enumsentence{\begin{tabular}[t]{@{}*{5}{l@{\ }}}
a. & Det \textbf{kommer att} bli mörkt snart & & b. &Jag \textbf{ska} gå och lägga mig nu\\
  & ``It will get dark soon" & &  &``I'm going to bed now"\\
\end{tabular}}
\enumsentence{
Jeg kommer til å savne deg ~~$[$NO$]$\\
``I will miss you"}
The modal verb \emph{`ska'} signals an intention of committing the action, either from the subject or from
the speaker. Cf.
\enumsentence{\begin{tabular}[t]{@{}*{5}{l@{\ }}}
a. & Du ska tycka om den & & b. &Du kommer tycka om den\\
  & ``You \emph{shall} like it" & &  &``You will like it."\\
\end{tabular}}
The verb \emph{`kommer'} (\emph{`come'}), normally used with the infinite marker \emph{att},
does not signal any intention, but that the speaker has belief that 
it will actually come true.\\
The resource grammar included 
\emph{`ska'}, which was implemented as the standard way of forming future
tense, and hence represents the translation of
\emph{`will'}. 
The new grammar also supports \emph{``kommer att"}, expressed as an alternative
future tense.
\begin{figure}[h]
\begin{verbatim}
(UseCl (TTAnt TFutKommer ASimul) PPos 
       (ImpersCl (AdvVP (ComplVA become_VA (PositA dark_A)) soon_Adv))
\end{verbatim}
\caption{Result of parsing ``Det kommer att bli mörkt snart". The future tense
         is marked by the constant TFutKommer}
  \label{fig:kommeratt}
\end{figure}

@@

The mapping %connects GF with another annotations and 
gives us means to evaluate our parser. By both parsing a Talbanken sentence and
transforming its annotated tree, we can easily inspect if the results are
equal.
Additionally, the mapping shows which grammatical constructions that are still missing
from the GF grammar and shows how the GF analysis differs from the one made
in Talbanken.
If there are words missing from our dictionary, the rich
POS-tags may help us to automatically find the correct declination and add it to the
lexicon. Further, our parser will need probabilities (see section
\ref{sec:future}) of how often a function is used. The GF treebank we
achieve from the translation is a good source for this information.\\
%missing words. Lexical aquisation, probabilities, see section .
%The mapping gives information about which form a word is
%currently used in and this may be used by the lexical extraction
%tools. % Those can be enhanced if they are given more data. 
%The translated trees enable us to extract those probabilities. %The mapping gives us a source for this.
%Furthermore, the translation makes it easy to identify grammatical constructions
%missing from the GF grammar Another important use of the mapping is evaluation of the parser, which can be
%accomplished by comparing the parse trees and the trees from the transformer.

@@

\noindent A similar issue arises from the use of verb particles. A verb with a
particle is treated as two independent units in Talbanken but in GF they are
considered to be one constituent. % which means that 
The particle has therefore no effect on the abstract tree, its presence is
announced by the verb itself.

@@

The flat version of Talbanken05 was used for developing the mapping, but
the program works for both the flat and the deep Tiger-XML versions.
The deeper contains six more tags, although they are not always useful for our       
purpose. Figure \ref{fig:mappDeepFlat} shows the difference for a conjunction.
% show piece of s5 here!
\begin{figure}[h]
\begin{parsetree}
    ( .CNP. 
        (.CJ.  (.NP. ~ `oskifta dödsbon'))
        ( .++. (.++OC. `och'))
        (.CJ.  (.VN. `familjestiftelser'))
    )
\end{parsetree}
\begin{parsetree}
    ( .CNP. 
        (.CJ.  (.NP. ~ `oskifta dödsbon'))
        ( .CC.
            (.CNP. 
                (.++.  (.++OC. `och' ))
                (.C+.  (.VN. `familjestiftelser'))
         ))
    )
\end{parsetree}
\caption{An example of the difference in the two analyses}
\label{fig:mappDeepFlat}
\end{figure}

Some more information about VP are given by the tag \verb|VG| (\emph{verb group}),
which groups an infinitival verb and its object together into a \verb|VP|.
This information can however be extracted from the flat implementation, and the results
get slightly better when using this version. \\

>>>>

\abstract{
This thesis describes work towards a wide-coverage grammar for
parsing and generating Swedish text.
We do this by using the dependently typed grammar formalism GF,
a functional programming language specialized at describing grammars.
The idea is to 
%The goal of this thesis is to enable open-domain parsing for Swedish
%by using the GF \\grammar formalism,
%by %The idea has been to% 
combine existing language resources with
new techniques, with an aim to achieve a parser
for unrestricted Swedish.
To reach this goal, problems of computational as well as linguistic nature had
to be solved.  
The work includes the development of the grammar -- \;
%in the GF formalism
to identify and formalize 
grammatical constructions %specific to and
frequent in Swedish -- as well as methods for importing a large-scale lexicon and for evaluating
the parser. % have been investigated and formalised in the functional
%programming language.
%~~achieve a covering Swedish~~ grammar implemented
%and the aim to reaching our goal;
%TODO first sentence should be more, no understatements!
%in GF -- to model natural language by functional programming. 
We present the methods and technologies used 
and discuss the advantages and problems of using GF
for modeling large-scale grammars.
We further discuss how our long-term goal can be reached
%the possibilities
%of reaching our long-term goal % -- to accomplish a parser for unrestricted Swedish -- 
by combining our rule-based grammar with statistical means.
%We compare and discuss .
%and for obtaining a system for analysing written text.
\\Our contribution is a wide-coverage GF lexicon, a translation of a Swedish treebank
into the GF notation and an extended Swedish grammar implementation. The grammar is based
on the multilingual abstract syntax given in the GF resource library, and now
also covers constructions specific to Swedish.
We further give an example of the advantage of using dependent types when
describing grammar and syntax, in this case for dealing with reflexive pronouns.
%the linguistic phenomena

@@

\newpage
\pagestyle{empty}
\section*{Acknowledgments}
\vspace{5mm}
Many people have helped me during this project and made this work possible.
I would first of all like to thank
Center of Language Technology, that has funded the project.

@@

The use of controlled language thus gives the possibility of good and reliable
translation and GF has successfully been used for this in a number of projects.
%like %Some examples of projects using GF are 
WebAlt \cite{webalt} aims to develop language independent
material for mathematics problems by using GF grammars.
The formal software verification tool
KeY \cite{key} used GF for translating formal specification to
natural language.
%>> key, talk, hats, bom (see references in Ramonas thesis).
%>> use talk here and move key to other place??
GF has further been used for describing dialogue grammars \cite{talk}, and
the framework is one of the main component in the
European project MOLTO\footnote{http://www.molto-project.eu/}, which develop
online translation between to 15 languages.\\

@@

  fun 
    Pred : N -> V -> S ;
    cat_N : N ;
    sleep_V : V ;
 }
\end{verbatim}
\caption{A small example of an abstract syntax}
\label{fig:gfAbstract1}
\end{figure}
\newpage % until behave

@@

\noindent We now have two implementations of the abstract, one for Swedish and one for English.
The resulting GF grammar is able both to
parse a string to an abstract tree and to go in the other direction; to produce
a string of natural language given an abstract tree. This step is called linearization.
Translation is a consequence of this, we can parse a Swedish string and then 
linearize the resulting abstract tree to English.

@@

\noindent The semantic aspect of Saldo requires that words with multiple meanings are
separated into different entries.
%TODO change word? uppskatta, känna, få
The word `uppskatta' for example, has two entries, with slightly
different semantics: \emph{enjoy}
\enumsentence{Jag uppskattar teater. \\
``I enjoy theater."}
\newpage % until behave
or \emph{estimate}: 
\enumsentence{Värdet uppskattades till 5 miljoner.\\
``The value was estimated to 5 million."}
%\enumsentence{Barnen växte varje dag\\
%``The children grew for each day"}
%or \emph{vegetate}. 
%\enumsentence{Det växte en gammal ek där.\\
%``There stood an old oak."}
`uppskatta' is however declined the same way for both interpretations of the word.
Hence \textit{Saldo's morphological lexicon} only keeps one entry of `uppskatta'.
For the homonyms that do not share the whole inflection tables, different entries are 
kept also in the morphological lexicon.\\
For our purpose, only the morphological section, provided under LPGL in XML
format, was needed.
The data can be processed and translated to GF format as described in 
section \ref{sec:prog.saldo}.

@@

%
%There are a number of differences between and saldo Till skillnad från GF,
%saldo does not contain forms that aren't used gf generates all forms. A form
%needed to generate the gf table may not be given in saldo such as singular
%forms for 'byxor'. Also contains idioms, 'hålla huvudet kallt' and has
%different grammatical analyse, a lot of pronouns. Example. 
\noindent Most but not all words of these categories have been imported.
One reason why the importing phase would fail 
is that Saldo, unlike GF, only contains the actually used word forms.
% does not contain information about ~~word forms that are not used~~.
For technical reasons, the smart paradigm might need forms never used.
%for this is Deponent verbs
%are for example given with their (often non-existing) active form, al
%In some cases those forms are needed by the smart paradigm. 
Consider for example
the plural tantum % \cite[44]{SAG} 
noun \emph{`glasögon'} (\emph{`glasses'}).
The smart paradigm requires a singular form, and since the program could not
find this in Saldo, there was no way of adding the lemma to the lexicon. 
When the program failed to import a noun, this was often the explanation.
%bbMany nouns fai, \emph{`kläder'} (\emph{`clothes'} etc.
Words of this type may be added manually, for \emph{`glasögon'} we could use
the ostensibly correct singular form `glasöga', although this
has another meaning (`glass-eye').
The same problem occurred for the irregular s-verbs,
(\emph{`synas'} \emph{(`show')} or \emph{umgås} \emph{(`socialize')})
which made up 61.5 \% of the failing verbs of type \verb_vb_.\\
%`frodas' (`flourish'), `lyckas' (`succeeed').
In a few cases the smart paradigms could not generate the correct declination.\\

@@

%       The table suggests that many 
%       >> 2537 without IDs and only the ones built of Alpha,only 552 without compounds, still contains abbrev.
%       >> 494 without abbr. spelling errors
%       >> 142 more then one time.
%       numerals and proper names, but some names are still there). In addition, there were 1817 words
%       >> 223 were IDS.++146 HS.
%       >> 426 more than one time, with compounds.
%       >> give reasons! should look good, not bad!
\noindent A list of words that were given different labels in GF than in Talbanken has been
composed, consisting of about 1600 entries. Many of those are
acceptable and reflects the difference 
made in the analyses, such as the examples in table \ref{tab:saldodiff}.
Others are examples of words that are still missing from
the lexicon.
\begin{figure}[h]
\begin{tabular}{|lll|}
\hline
Word    & Talbanken tag & GF category \\
\hline
måste   & MVPS          & VV \\
allting & POTP          & N \\
få      & POZP          & Det \\
\hline
\end{tabular}
\caption{}
\label{tab:saldodiff}
\end{figure}

%Since Saldo does not give any valency information, neither does the imported lexicon.
%The lexicon does not contain any information.
\noindent Valency information, which is crucial for GF, is not given in Saldo and
hence not in the imported lexicon. It remains as future work to find methods 
%using the results from the mapping (see section \ref{sec:Mapping})
to extract this information from Talbanken and to automatically build
it into the lexicon. 
%The saddest thing: no valency info. Crucial for GF. Only reflexive and verbs with particles
%have any information. The earlier explained techniques may be used for this in Talbanken.
%To get even bigger material, use Korp, but then no guarantee.

@@

\begin{figure}[h]
\begin{verbatim}
DetCN : Det                  -> CN     -> NP ;
        indefinite, singular -> flicka -> "en flicka"
        definitive, singular -> flicka -> "flickan"
\end{verbatim}        
\caption{}
\label{gfcode:DetCN}
\end{figure}
\newpage
%The two basic articles are \verb-IndefArt- and \verb-DefArt- which
%uses the post-modified form of the noun: katten.
\noindent Some rules for determination were described in section \ref{sec:swedishnoun};
if we have a common noun phrase consisting of the parts \emph{`liten'}
(\emph{`small'}) and \emph{`katt'} (\emph{`cat'}), there are three ways they
can be combined, as shown in figure \ref{tab:detdef}. The noun may be used in
definite or indefinite form and the adjective in its weak or strong form \cite[p. 31]{gunlog}.
\begin{figure}[h]
\begin{tabular}{|lll|l|}
\hline
\textbf{Determiner} & \textbf{Adjective} & \textbf{Noun} & \textbf{DetSpecies}\\
\hline
en  &  liten $[${\sc -Def}$]$& katt   $[${\sc -Det}$]$& \textbf{ DIndef}\\
min &  lilla $[${\sc +Def}$]$& katt   $[${\sc -Det}$]$  & \textbf{ DDef Indef} \\
den &  lilla $[${\sc +Def}$]$& katten $[${\sc +Det}$]$& \textbf{ DDef Def} \\
\hline
\end{tabular}        
\caption{}\label{tab:detdef}
\end{figure}

@@

\noindent Swedish verbs may take up to five arguments \cite[p. 53]{stymne}.
These may be prepositions, particles, reflexive object, indirect objects
and direct objects. 
\enumsentence{\shortex{7}
{Jag &tar& med &mig& den& till& honom}
{\emph{I} &\emph{take}&\emph{with} &\emph{me}& \emph{it}& \emph{to}& \emph{him}}
{``I bring it to him"}
\label{sent:tamed}}
The verb `ta' in sentence (\ref{sent:tamed}) takes one particle (\emph{`med'}),
one preposition (\emph{`till'}) and two objects: \emph{`den'} and \emph{`honom'}.
In a GF lexicon this verb is given the
category \verb-V3-, %classifing it as 
a \textit{three-place verb}, taking two objects.
The notion \verb-V3- is motivated by the formal translation: \verb-bring(I,it,him)-.\\
%these will be encoded in different ways.
Particles are given in the lexicon, as well as the prepositions that are
chosen by the verb, since these may not vary.
The entry for `\emph{ta med}', as used in sentence (\ref{sent:tamed}), is
described as follows:
%, assuming that we already have an constant \verb-take_V- which produces the correct conjugation for `ta':
\begin{verbatim}
ta_med_V3 = dirV3 (reflV (partV (take_V "med"))) (mkPrep "till") ;
\end{verbatim}
%The category \verb-V3- tells us that the verb takes two objects, % (`den' and `honom').
The function \verb-dirV3- creates a three-place verb, where the first object is direct and
the second is to be used with the preposition given as the last argument: `\emph{till}'.
\verb-reflV- shows that the verb always is used with a reflexive
pronoun and \verb-partV- gives the particle \emph{`med'}.
The fact that the chosen prepositions is attached to the verb in the lexicon
causes the parse tree visualization algorithm to group them together.
This is also the case for particles, cf. parse tree \ref{pic:tittapart}
and \ref{pic:tittapa}.
\begin{figure}[h]
\subfloat[Verb with a chosen preposition]{\label{pic:tittapa}\includegraphics[height=70mm]{tittapa.jpg}}
\hspace{10mm}
\subfloat[Verb with particle]{\label{pic:tittapart}\includegraphics[height=76mm]{tittapart.pdf}}
\caption{The visualized parse trees do not show the internal difference of chosen prepositions and particles}
\label{fig:translationtrees}
\end{figure}
\newpage
\noindent As already stated, the visualized parse trees is not a complete representation,
even if the verb phrases in the visualizations look the same, the two cases
are treated and represented differently internally. The fronting of the
preposition, as in of sentence (\ref{sent:tittapa}a)., is accepted but
fronting of particles, as in \ref{sent:tittapa}b., is not.
%TODO use png to get correct size
\eenumsentence{
\item[a]\shortex{4}
{På &pojken& tittar& du.}
{\emph{on} &\emph{the boy}& \emph{look}& \emph{you}}
{``You look at the boy."}
\item[b]\shortexnt{6}
{*På& när& du& springer& tittar& jag}
{\emph{~on}& \emph{when}& \emph{you}& \emph{run}& \emph{watch}& \emph{I}}
\label{sent:tittapa}}

@@

\noindent Like verb phrases, a clause may also be missing an object and then has the type
\verb-ClSlash-.
The \verb-ClSlash- is formed by a \verb-VPSlash- which is given a subject.
This is a convenient way to form questions, relative clauses
and topicalized clauses (see figure \ref{fig:clause2}), as introduced in \cite{gazdar}.\\

@@

%What was in the resources
%description of the existing grammar, the types, the possibilities
%VP, NP.
%valency built in (problems with ställa - på, under, vid).
%Variants, nonexist.
%coverage of standard Swedish.
\newpage % until behave
\section{Development of the grammar}
\label{sec:Added}
%The resource grammar is meant to cover the most important grammatical structures
%present in many languages, but not the language specific constructions.
It has earlier been hard to identify missing constructions of the Swedish
implementation, since there was no large resource available to evaluate it on.

@@

%'han såg sin mammas bil', they same could be applied to
%APs for example \ref{ex:ReflAP} or Adv for
%\enumsentence{Han leker oftare än sin kompis}
% (ComparAdvAdj)
%Atm you can't focus this kind of object, then a ClSlash need info about the
%subject. FunRP also tricky (and what does it mean, we can say
%Han såg sin katt i vilken musen låg), in this case RP needs to give more information.
%Kort sagt, the information about the subject must be spread
%in different parts of the sentence. 
%Could also decide to always produc 'sin' -> jag såg sin bil
%to avoid the dependence, but not nice. Or keep a flag in NP saying if it is
%in third person, and add nonexist for others. But this way we can also get rid
%of ReflVP and the possessive prons \verb|PossPron| should only be used for somebody else:
%'jag såg din bil'.
%Unfortunaly, have to change ReflVP, this only allows this in object for VPSlash, not FocObj,
%not 'alla sina', usw. Want it to handle conjuction as in example \ref{ex:refGen}
%etc.
%varandra, works the same but has no case for Sg.
%%
\newpage %until behave
\subsection{A second future tense: ``Kommer att"}
Swedish and Norwegian use two auxiliary verbs for future tense \\
\cite[p. 246]{H&H}:
%\cite[Tempus \textsection 28]{SAG}.
\enumsentence{\begin{tabular}[t]{@{}*{5}{l@{\ }}}
a. & Det \textbf{kommer att} bli mörkt snart & & b. &Jag \textbf{ska} gå och lägga mig nu\\
  & ``It will get dark soon" & &  &``I'm going to bed now"\\
\end{tabular}}
\enumsentence{
Jeg kommer til å savne deg ~~$[$NO$]$\\
``I will miss you"}
The modal verb \emph{`ska'} signals an intention of committing the action, either from the subject or from
the speaker. Cf.
\enumsentence{\begin{tabular}[t]{@{}*{5}{l@{\ }}}
a. & Du ska tycka om den & & b. &Du kommer tycka om den\\
  & ``You \emph{shall} like it" & &  &``You will like it."\\
\end{tabular}}
The verb \emph{`kommer'} (\emph{`come'}), normally used with the infinite marker \emph{att},
does not signal any intention, but that the speaker has belief that 
it will actually come true.\\
The resource grammar included 
\emph{`ska'}, which was implemented as the standard way of forming future
tense, and hence represents the translation of
\emph{`will'}. 
The new grammar also supports \emph{``kommer att"}, expressed as an alternative
future tense.
\begin{figure}[h]
\begin{verbatim}
(UseCl (TTAnt TFutKommer ASimul) PPos 
       (ImpersCl (AdvVP (ComplVA become_VA (PositA dark_A)) soon_Adv))
\end{verbatim}
\caption{Result of parsing ``Det kommer att bli mörkt snart". The future tense
         is marked by the constant TFutKommer}
  \label{fig:kommeratt}
\end{figure}

@@

The translation %connects GF with another annotations and 
gives us means to evaluate our parser. By both parsing a Talbanken sentence and
transforming its annotated tree, we can easily inspect if the results are
equal.
Additionally, the mapping shows which grammatical constructions that are still missing
from the GF grammar and shows how the GF analysis differs from the one made
in Talbanken.
If there are words missing from our dictionary, the rich
POS-tags may help us to automatically find the correct declination and add it to the
lexicon. Further, our parser will need probabilities (see section
\ref{sec:future}) of how often a function is used. The GF treebank we
achieve from the translation is a good source for this information.\\
%missing words. Lexical aquisation, probabilities, see section .
%The mapping gives information about which form a word is
%currently used in and this may be used by the lexical extraction
%tools. % Those can be enhanced if they are given more data. 
%The translated trees enable us to extract those probabilities. %The mapping gives us a source for this.
%Furthermore, the translation makes it easy to identify grammatical constructions
%missing from the GF grammar Another important use of the mapping is evaluation of the parser, which can be
%accomplished by comparing the parse trees and the trees from the transformer.

@@

\noindent A similar issue arises from the use of verb particles. A verb with a
particle is treated as two independent units in Talbanken but in GF they are
considered to be one constituent. % which means that 
The particle has therefore no effect on the abstract tree, its presence is
announced by the verb \\itself.

@@

The flat version of Talbanken05 was used for developing the mapping, but
the program works for both the flat and the deep Tiger-XML versions.
The deeper contains six more tags, although they are not always useful for our       
purpose. Figure \ref{fig:mappDeepFlat} shows the difference for a conjunction.
% show piece of s5 here!
\begin{figure}[h]
\begin{parsetree}
    ( .CNP. 
        (.CJ.  (.NP. ~ `oskifta dödsbon'))
        ( .++. (.++OC. `och'))
        (.CJ.  (.VN. `familjestiftelser'))
    )
\end{parsetree}
\begin{parsetree}
    ( .CNP. 
        (.CJ.  (.NP. ~ `oskifta dödsbon'))
        ( .CC.
            (.CNP. 
                (.++.  (.++OC. `och' ))
                (.C+.  (.VN. `familjestiftelser'))
         ))
    )
\end{parsetree}
\caption{An example of the difference in the two analyses}
\label{fig:mappDeepFlat}
\end{figure}
The focus on the mapping has been simple sentences, in order to cover the most
fundamental structures. If the program is extended and adapted to cover complex
conjunction, a deeper evaluation should be preformed to see which version that
suits our needs the best.
The deeper implementation also gives 
some more information about verb phrases, by the tag \verb|VG| (\emph{verb group}),
which groups an infinitival verb and its object together into a \verb|VP|.
This information can however be extracted from the flat implementation, and the results
get slightly better when using this version. \\
